---
title: "Uncovering non-monotonic dependence with aceshuffle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncovering non-monotonic dependence with aceshuffle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 3.9,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r library}
library(udp)
```

```{r guard, include = FALSE}
knitr::opts_chunk$set(eval = requireNamespace("copula", quietly = TRUE))
```

Stochastic inversion (see `vignette("udp")`) lets us build a dataset with a
deliberately awkward dependence structure. Start from a strongly dependent
copula sample and pull each margin back through a udp function: the margins
stay uniform, but their ordinary correlation can be made to vanish even
though the pair is far from independent. This vignette uses the **copula**
package to build such a pair, then uses `aceshuffle()` to recover the
non-monotonic dependence.

```{r data}
set.seed(1)
n <- 1000

udp2 <- udpcosine(6)
# odd degree, unlike udp2: its shape isn't close to symmetric about the centre
# of the unit square, which matters below when aceshuffle() has to pick an
# orientation
udp1 <- udplegendre(5)

GC <- copula::gumbelCopula(3)
W <- copula::rCopula(n, GC)
X <- cbind(udpsi(udp1, W[, 1]), udpsi(udp2, W[, 2]))
plot(X, pch = 20, cex = 0.4, xlab = "X1", ylab = "X2", asp = 1, xaxs = "i", yaxs = "i")
```

The margins of `X` are uniform and their ordinary correlation is close to zero,
yet they are far from independent. `aceshuffle()` searches, in the manner of
alternating conditional expectations, for a shuffle of each margin that makes
the transformed pair as linearly correlated as possible.

```{r ace-fit}
fit <- aceshuffle(X, m = 50)
fit@iterations
fit@correlation
```

`aceshuffle()` returns a \linkS4class{bishuffle} object, packaging the two
fitted shuffles together with a `plot()` method that draws them side by side:

```{r ace-plot, fig.width = 7.8}
plot(fit, embellish = "colour")
```

The fitted shuffles approximate the udp functions we used to build the data.
The fitted pair is unique only up to the joint reflection
`(shuffle1, shuffle2) -> (1 - shuffle1, 1 - shuffle2)`, which leaves the
correlation unchanged; `aceshuffle()` returns the orientation whose shuffles
trend upward.

```{r ace-shapes, fig.show = "hold", fig.width = 3.9}
u <- seq(0, 1, length.out = 200)

plot(fit@shuffle1, embellish = "colour")
lines(u, udptrans(udp1, u), col = 2, lwd = 2)

plot(fit@shuffle2, embellish = "colour")
lines(u, udptrans(udp2, u), col = 2, lwd = 2)
```

When a generating transform is close to symmetric about the centre of the unit
square (as `udpcosine(6)` is), the trend rule cannot pin the orientation from
that shuffle alone; here the clear upward trend of the `udplegendre(5)` shuffle
fixes the orientation of the pair. If an overlay comes out reflected, compare
against `1 - udptrans(...)` instead.

`udptrans(fit, X)` applies the two shuffles to their matching columns,
`cbind(udptrans(shuffle1, X[, 1]), udptrans(shuffle2, X[, 2]))`. Its
correlation matches the copula it came from.

```{r ace-result}
V <- udptrans(fit, X)
plot(V, pch = 20, cex = 0.4, asp = 1, xaxs = "i", yaxs = "i")
c(fitted = fit@correlation, original = cor(W)[1, 2])
```

## References

- McNeil, A. J., Nešlehová, J. G. and Smith, A. D. (2025). Measures and
  models of non-monotonic dependence. arXiv:2512.10828.
