---
title: "Bivariate stochastic inversion copulas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bivariate stochastic inversion copulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 3.9,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r library}
library(udp)
```

```{r guard, include = FALSE}
knitr::opts_chunk$set(eval = requireNamespace("rvinecopulib", quietly = TRUE))
```

A `bsicopula` couples two `udp` transformations, `udp1` and `udp2`,
through a copula on their "carrier" uniforms `(V1, V2)`. Margin `i`'s draw
is `udpsi(udp_i, V_i, Z_i)`: `udpsi()` applied to `V_i`, with the
pre-image selected by a randomizer `Z_i`. With independent randomizers,
that pre-image selection is the only source of randomness beyond
`basecopula` itself; a **randomizer model** -- an `sdvine` or a
`vmixture` -- instead couples `Z1` and `Z2` (on `(V1, V2)` and on each
other), injecting further dependence between the two margins.

This vignette builds four `bsicopula` objects that share the same
`basecopula` and the same margins (both `vsymmetric()`, the two-branch fold
`T(u) = |2u - 1|`), differing only in their randomizer model, and finishes
by uncovering something all four have in common despite looking
completely different.

```{r library2}
library(rvinecopulib)
```

## Four ways to couple two v-transforms

### Independent randomizers

With `randomizermod` left at its default `NULL`, each margin's pre-image is
selected independently -- the joint density is exactly `c_V(T1(u1),
T2(u2))` for `c_V` the density of `basecopula` (see `dbsicopula()`).

```{r case1}
case1 <- bsicopula(bicop_dist("gauss", parameters = 0.85),
  udp1 = vsymmetric(),
  udp2 = vsymmetric()
)
plot(case1)
```

The four-petal shape is entirely inherited from `vsymmetric()` folding a
Gaussian copula's own elliptical contours onto themselves; there is no
extra coupling between `Z1` and `Z2` yet.

### A single extra dependency

An `sdvine` model's `copZ1Z2_V1V2` slot (its only required argument)
couples `Z1` and `Z2` directly, given `(V1, V2)`.

```{r case2}
ranmodA <- sdvine(bicop_dist("gauss", parameters = 0.8))
case2 <- bsicopula(bicop_dist("gauss", parameters = 0.85),
  udp1 = vsymmetric(),
  udp2 = vsymmetric(),
  randomizermod = ranmodA
)
plot(case2)
```

The central peak stretches along the main diagonal: with `Z1` and `Z2`
now positively dependent, the same branch tends to get selected for both
margins together, on top of the folding structure already present in
`case1`.

### A fuller D-vine

`sdvine`'s other two slots add the remaining tree-2 edges, `Z1`-`V2` given
`V1` and `V1`-`Z2` given `V2`, coupling each randomizer to the *other*
margin's carrier uniform as well.

```{r case3}
ranmodB <- sdvine(
  bicop_dist("gauss", parameters = 0.8),
  bicop_dist("gauss", parameters = 0.7),
  bicop_dist("gauss", parameters = 0.1)
)
case3 <- bsicopula(bicop_dist("gauss", parameters = 0.85),
  udp1 = vsymmetric(),
  udp2 = vsymmetric(),
  randomizermod = ranmodB
)
plot(case3)
```

### Regime-switching dependence

A `vmixture` model takes a different approach: rather than
coupling `Z1` and `Z2` smoothly, it switches between two copulas
altogether, chosen by a `selector(v1, v2)`. Here `Z1` and `Z2` are drawn
comonotonically when `V1` or `V2` is large, and countermonotonically
otherwise.

```{r case4}
ranmodC <- vmixture(
  bicop_dist("gauss", parameters = 1),
  bicop_dist("gauss", parameters = -1),
  selector = function(v1, v2) pmax(v1, v2) > 0.6
)
case4 <- bsicopula(bicop_dist("gauss", parameters = 0.85),
  udp1 = vsymmetric(),
  udp2 = vsymmetric(),
  randomizermod = ranmodC
)
plot(case4)
```

The abrupt regime switch leaves a visible seam in the density where
`pmax(v1, v2)` crosses `0.6`, quite unlike the smooth reshaping in `case2`
and `case3`.

## An invariant hiding underneath

These four densities look nothing alike, yet they share something exactly,
not just approximately. `udpsi()` always returns a genuine pre-image of
its input -- that is the whole point of stochastic inversion -- so
`udptrans(udp_i, udpsi(udp_i, V_i, Z_i))` equals `V_i` *identically*,
whatever `Z_i` turned out to be, chosen by any randomizer model at all
(or none). Applying `udptrans(vsymmetric(), .)` to each margin of a
`bsicopula` sample is therefore not a statistical recovery of `(V1, V2)`;
it is an exact algebraic one, for every single draw.

```{r invariant-exact}
set.seed(1)
V <- rbicop(10, case1@basecopula)
Z1 <- runif(10) # any randomizer at all -- not even a valid one matters here
Z2 <- runif(10)
U1 <- udpsi(vsymmetric(), V[, 1], Z1)
U2 <- udpsi(vsymmetric(), V[, 2], Z2)
max(abs(udptrans(vsymmetric(), U1) - V[, 1]))
max(abs(udptrans(vsymmetric(), U2) - V[, 2]))
```

Since `basecopula` is the *same* Gaussian copula in all four cases, this
means `(udptrans(vsymmetric(), U1), udptrans(vsymmetric(), U2))` has
exactly the same distribution for every one of them -- the correlation
`0.85` of the shared `basecopula` -- no matter how different the raw
`(U1, U2)` densities plotted above look.

```{r invariant-distributional}
set.seed(1)
n <- 20000
for (nm in c("case1", "case2", "case3", "case4")) {
  bc <- get(nm)
  U <- rbsicopula(n, bc)
  V_recovered <- cbind(udptrans(vsymmetric(), U[, "U1"]), udptrans(vsymmetric(), U[, "U2"]))
  cat(nm, ": cor(recovered V) =", cor(V_recovered)[1, 2], "\n")
}
```

In other words, all four `bsicopula` objects here are different
*refinements* of the very same coarse dependence structure on `(V1, V2)`.
A randomizer model never touches that shared structure -- it only decides
*which* pre-image `vsymmetric()`'s fold selects for each margin, and that
selection is entirely where the visibly different dependence patterns
above come from: linear correlation the base copula already carries in
`case1`, extra diagonal stretching in `case2` and `case3`, or an outright
regime switch in `case4`. Coarsen any of them back through `vsymmetric()`
and that difference disappears completely.

## References

McNeil, A. J. and Nešlehová, J. G. (2026). Stochastic inversion of
multivariate uniform-distribution-preserving transformations.
arXiv:2607.07174.
