---
title: "Fitting bivariate stochastic inversion copulas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting bivariate stochastic inversion copulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.8,
  fig.align = "center"
)
```

```{r guard, include = FALSE}
knitr::opts_chunk$set(eval = requireNamespace("rvinecopulib", quietly = TRUE))
```

```{r library}
library(udp)
library(rvinecopulib)
```

`fitbsicopula()` fits a `bsicopula` model to copula data by maximum
likelihood. The object you give it fixes the *structure* of the model -- the
family and rotation of the base copula, the class of each udp
transformation, and the kind of randomizer model -- and supplies the
*starting values* of the parameters. Everything continuous is then estimated.
This vignette fits three models to simulated data, in each case starting
well away from the truth:

1. independent randomizers, with a three-piece zigzag and a `v2p`
   v-transform as the udp transformations;
2. a `randsdvine` randomizer model, with v-transforms for the udps;
3. independent randomizers and the absolute spherical t copula as the
   base copula, with v-transforms for the udps.

A third example uses the absolute spherical t copula as the base copula. Each follows the same recipe: simulate from a known model, estimate it from
poor starting values, and compare the true and fitted densities.

The data given to `fitbsicopula()` are pseudo-observations, ranks scaled into
`(0, 1)`, as is usual for copula estimation. In practice they would come from
a data set through `pseudo_obs()`; here we simulate copula data and take the
ranks.

## Example 1: independent randomizers, a zigzag and a v-transform

### The true model and the data

The first margin uses `udpzigzag()`, a piecewise-linear transformation
whose pieces alternately rise and fall; here it has three pieces, of widths
`2/7`, `3/7` and `2/7`. The second uses `v2p()`, a smooth v-transform with
fulcrum `delta` and shape `kappa`. The base copula is Gaussian.

```{r truth1, fig.width = 3.9, fig.height = 4.5}
truth1 <- bsicopula(bicop_dist("gaussian", 0, 0.7),
  udp1 = udpzigzag(widths = c(2, 3, 2)),
  udp2 = v2p(delta = 0.35, kappa = 1.5)
)
set.seed(11)
U1 <- pseudo_obs(rbsicopula(1000, truth1))
plot(U1, pch = 20, cex = 0.5, col = "grey35", xlab = "u1", ylab = "u2", asp = 1, xaxs = "i", yaxs = "i")
```

### Fitting from poor starting values

The starting object has the same structure but a weaker correlation, a
zigzag of three equal pieces, and a v-transform with the fulcrum in the
middle and no extra curvature.

```{r start1}
start1 <- bsicopula(bicop_dist("gaussian", 0, 0.3),
  udp1 = udpzigzag(widths = c(1, 1, 1)),
  udp2 = v2p(delta = 0.5, kappa = 1)
)
fit1 <- fitbsicopula(U1, start1)
fit1
```

The parameters are reported on their natural scale: `rho` for the Gaussian
copula, the two interior breakpoints of the zigzag (the true ones are
`2/7 = 0.286` and `5/7 = 0.714`) and `delta` and `kappa` for the
v-transform. The number of pieces of the zigzag and its direction (`up`) are
part of the structure and come from the starting object.

Standard errors come from a parametric bootstrap: new samples are drawn from
the fitted model, ranked, and put through the whole procedure again. It
takes a little while, so we use a modest number of replicates here.

```{r boot1, cache = FALSE}
set.seed(1)
fit1se <- fitbsicopula(U1, start1, se = "bootstrap", B = 40)
true1 <- c(0.7, 2 / 7, 5 / 7, 0.35, 1.5)
cbind(truth = true1, signif(cbind(estimate = coef(fit1se), se = fit1se@se), 3))
```

### True and fitted densities

The fitted model is the `bsicopula` stored in the fit.

```{r dens1, fig.height = 4.2}
fitted1 <- fit1@bsicopula
op <- par(mfrow = c(1, 2))
plot(truth1, main = "True density")
plot(fitted1, main = "Fitted density")
par(op)
```

### h-functions and the joint CDF

Since the fitted object is an ordinary `bsicopula`, the other functions of
the package apply to it. `hbsicopula()` gives the conditional distribution
function `h(u2 | u1) = P(U2 <= u2 | U1 = u1)`. Because the second margin is a
smooth v-transform rather than a piecewise-linear one, `hbsicopula()`
computes it by numerical quadrature (with piecewise-linear transformations
and independent randomizers it is an exact finite sum); the true and fitted
models are equally easy to evaluate.

```{r h1}
u2 <- seq(0, 1, length.out = 201)
u1s <- c(0.15, 0.5, 0.85)
plot(NA, xlim = c(0, 1), ylim = c(0, 1), xlab = "u2", ylab = "h(u2 | u1)")
abline(0, 1, col = "grey85")
for (k in seq_along(u1s)) {
  lines(u2, hbsicopula(u1s[k], u2, truth1), col = k + 1, lwd = 2)
  lines(u2, hbsicopula(u1s[k], u2, fitted1), col = k + 1, lwd = 2, lty = 2)
}
legend("bottomright", c(paste("u1 =", u1s), "true", "fitted"),
  col = c(2:4, 1, 1), lwd = 2, lty = c(1, 1, 1, 1, 2), bty = "n"
)
```

Conditional on `u1`, the curves rise and fall as the zigzag sends `u1` to
different parts of the carrier scale. The joint distribution function
`pbsicopula()` is compared in the same way, through the departure from
independence `C(u1, u2) - u1 u2`.

```{r cdf1, fig.height = 4.2}
g <- (1:40 - 0.5) / 40
G <- expand.grid(u1 = g, u2 = g)
dep <- function(m) matrix(pbsicopula(G$u1, G$u2, m, nodes = 41) - G$u1 * G$u2, 40)
Ct <- dep(truth1)
Cf <- dep(fitted1)
lev <- pretty(range(Ct, Cf), 10)
op <- par(mfrow = c(1, 2))
contour(g, g, Ct, levels = lev, main = "True C(u1, u2) - u1 u2", xlab = "u1", ylab = "u2")
contour(g, g, Cf, levels = lev, main = "Fitted", xlab = "u1", ylab = "u2")
par(op)
max(abs(Ct - Cf))
```

## Example 2: a `randsdvine` randomizer model with v-transforms

### The true model and the data

In a `bsicopula` with a randomizer model the randomizers `Z1` and `Z2`,
which select the branch of each udp transformation, are no longer
independent. `randsdvine()` couples them through a D-vine on
`(Z1, V1, V2, Z2)`; by default only the tree-3 copula `copZ1Z2_V1V2`,
the dependence of `Z1` and `Z2` given the carrier values, is non-trivial.
Here the base copula is Gumbel, which has upper tail dependence, and the
margins are a `v2p` and a `vlinear` v-transform.

```{r truth2, fig.width = 3.9, fig.height = 4.5}
truth2 <- bsicopula(bicop_dist("gumbel", 0, 2),
  udp1 = v2p(delta = 0.4, kappa = 1.5),
  udp2 = vlinear(delta = 0.6),
  randomizermod = randsdvine(bicop_dist("gaussian", 0, 0.6))
)
set.seed(12)
U2 <- pseudo_obs(rbsicopula(800, truth2))
plot(U2, pch = 20, cex = 0.5, col = "grey35", xlab = "u1", ylab = "u2", asp = 1, xaxs = "i", yaxs = "i")
```

### Fitting

The starting object carries the structure: the Gumbel family, the two
v-transform classes, and a `randsdvine` whose single Gaussian copula starts
near independence. The starting parameters are all far from the truth.

```{r start2}
start2 <- bsicopula(bicop_dist("gumbel", 0, 1.4),
  udp1 = v2p(delta = 0.5, kappa = 1),
  udp2 = vlinear(delta = 0.5),
  randomizermod = randsdvine(bicop_dist("gaussian", 0, 0.1))
)
fit2 <- fitbsicopula(U2, start2)
fit2
```

By default `twostage = TRUE`: the model is first fitted with independent
randomizers, and that fit supplies the starting values of the full model, so
the randomizer's copula is estimated last. The parameter `copZ1Z2_V1V2.rho`
is the correlation of the Gaussian copula coupling `Z1` and `Z2`.

Is the randomizer model needed? The same data fitted with independent
randomizers gives a markedly lower likelihood, and information criteria
favour the full model:

```{r indep2}
fit2null <- fitbsicopula(U2, bsicopula(start2@basecopula, start2@udp1, start2@udp2))
AIC(fit2null, fit2)
```

### True and fitted densities

```{r dens2, fig.height = 4.2}
op <- par(mfrow = c(1, 2))
plot(truth2, main = "True density")
plot(fit2@bsicopula, main = "Fitted density")
par(op)
```

Even though the model is fitted to ranks only, the fitted density reproduces
the main features of the true one: the shape imposed by the
v-transforms and the concentration in the upper-right corner from the
Gumbel copula's tail dependence.

## Example 3: the absolute spherical t copula as base copula

The base copula need not come from `rvinecopulib`. The absolute spherical
t copula, `astcopula(nu)`, is the copula of the absolute values of a
bivariate t vector with correlation zero (Dias, Han and McNeil, 2027). It
has upper tail dependence, a single parameter `nu`, and is available as a
base copula of a `bsicopula`. It interpolates between independence
(`nu` large) and comonotonicity (`nu` small), and its Kendall's tau is
`astcopula_tau(nu)`.

```{r truth3}
truth3 <- bsicopula(astcopula(2),
  udp1 = vlinear(delta = 0.4),
  udp2 = v2p(delta = 0.6, kappa = 1.4)
)
set.seed(21)
U3 <- pseudo_obs(rbsicopula(1500, truth3))
```

The fit starts from 15 degrees of freedom -- nearly independence -- and
fulcrums at the middle of the unit interval. The degrees of freedom are
estimated on the log scale and reported as `basecopula.nu`.

```{r fit3}
start3 <- bsicopula(astcopula(15),
  udp1 = vlinear(delta = 0.5),
  udp2 = v2p(delta = 0.5, kappa = 1)
)
fit3 <- fitbsicopula(U3, start3)
fit3
```

The fitted `nu` can be reported as a Kendall's tau of the base copula,
and `astcopula_nu()` converts a Kendall's tau into degrees of freedom:

```{r tau3}
c(true = astcopula_tau(2), fitted = astcopula_tau(coef(fit3)[["basecopula.nu"]]))
astcopula_nu(0.2)
```

```{r dens3, fig.height = 4.2}
op <- par(mfrow = c(1, 2))
plot(truth3, main = "True density")
plot(fit3@bsicopula, main = "Fitted density")
par(op)
```

The shape of this density comes from the two v-transforms, which fold the
carrier values of the t copula with correlation zero, so that ranks near
the fulcrums are pushed to the corner where the ast copula has its
tail dependence. As in Example 1, `se = "bootstrap"` would give standard
errors.

## Practical notes

* `fitbsicopula()` fits `bicop_dist` base copulas from `rvinecopulib` and
  the absolute spherical t copula `astcopula()`, and the randomizer model
  must be `NULL` or a `randsdvine`. (Example 3 uses the second.)
* The estimation of models with a `randsdvine` randomizer is harder than
  with independent randomizers: the likelihood can have several local
  maxima, which is why `fitbsicopula()` also starts the full fit from the
  starting values you supply and keeps the better of the two fits. Treat the
  fitted randomizer parameters with some caution and compare fits from
  different starting values.
* A udp whose parameters should not be estimated (determined elsewhere, for
  instance by `aceshuffle()`) is held fixed with `udpfix = TRUE`.
* The likelihood in a udp breakpoint such as `delta` is only piecewise
  smooth, with a fine sawtooth for many base copulas. The fit is therefore
  done by continuation, smoothing the carrier values first, and the
  breakpoint estimates are accurate to about `1 / n`. See `?fitbsicopula` for
  the details, and for why `se = "hessian"` should not be trusted for
  breakpoints.
* Set `pseudo = FALSE` when the data are probability integral transforms
  from fitted marginal models rather than ranks; this only changes the
  bootstrap.

## References

Dias, A., Han, J. and McNeil, A. J. (2027). GARCH copulas, v-transforms and
D-vines for stochastic volatility. *Journal of Multivariate Analysis*, **217**,
105695. <https://doi.org/10.1016/j.jmva.2026.105695>
