---
title: "Finite orthogonal-series udp transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finite orthogonal-series udp transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 3.9,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r library}
library(udp)
```

`udplegendre(j)` and `udpcosine(j)` each build a udp transformation from a
single degree-`j` term of an orthonormal basis on `[0, 1]` -- shifted
Legendre polynomials for one, cosines for the other. `udplegendrebex()` and
`udpcosinebex()` generalize both to a finite weighted *sum* of terms,
`g(u) = sum_j coef[j] * basis_j(u)`. The construction is otherwise the same
as for the single-term classes -- `T(u) = F(g(u))` for `F` the distribution
function of `g(U)` -- but a sum can have as many turning points as its
effective degree allows, and its range need not be symmetric, giving access
to shapes a single term cannot produce.

Throughout we work with a single uniform sample.

```{r data}
set.seed(1)
n <- 1000
U <- runif(n)
```

## A sum of Legendre polynomials

`udplegendrebex()` takes a vector of weights on the orthonormal shifted
Legendre polynomials. A one-term vector reproduces `udplegendre()` exactly;
combining several terms produces a curve with more turning points and a less
regular shape than any single degree could.

```{r legendre-sum}
T1 <- udplegendrebex(c(0.44, -0.35, 0.56, -0.21, 0.36))
plot(T1, embellish = "colour")
```

`udptrans()` applies the map; the transformed sample is still uniform.

```{r legendre-sum-forward}
V <- udptrans(T1, U)
hist(V, main = "udptrans(T1, U)")
```

The map is many-to-one, so it has no ordinary inverse. `udpsi()` draws one
pre-image at random, in the proportion that keeps the uniform property
running backwards.

```{r legendre-sum-inverse}
U2 <- udpsi(T1, V)
hist(U2, main = "udpsi(T1, V)")
```

`pcoincide()` gives the probability that stochastic inversion lands back on
the original value.

```{r legendre-sum-pcoincide}
plot(U, U2, pch = 20, cex = 0.4, asp = 1, xaxs = "i", yaxs = "i")
pcoincide(T1)
```

## A sum of cosine functions

`udpcosinebex()` is built the same way from the orthonormal cosine
functions. Internally it works with the Chebyshev polynomial that
`g(u) = h(cos(pi * u))` reduces to, but the interface -- construction, plot,
`udptrans()`, `udpsi()`, `pcoincide()` -- is identical to
`udplegendrebex()`.

```{r cosine-sum}
T2 <- udpcosinebex(c(0.44, -0.35, 0.56, -0.21, 0.36))
plot(T2, embellish = "colour")

V <- udptrans(T2, U)
U2 <- udpsi(T2, V)
pcoincide(T2)
```

## The same weights, two different shapes

Passed the same coefficients, the two constructions give visibly different
curves -- one built from a polynomial, the other from a trigonometric sum --
despite sharing an orthonormal-expansion origin, a stochastic inverse, and
the uniform-distribution-preserving property itself.

```{r compare, fig.show = "hold", fig.width = 3.9}
plot(T1, main = "udplegendrebex", embellish = "colour")
plot(T2, main = "udpcosinebex", embellish = "colour")
```

## References

- McNeil, A. J., Nešlehová, J. G. and Smith, A. D. (2025). Measures and
  models of non-monotonic dependence. arXiv:2512.10828.
