---
title: "Uniform distribution preserving transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uniform distribution preserving transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 3.9,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r library}
library(udp)
```

A map `T` of the unit interval to itself is **uniform distribution preserving**
(udp) if `T(U)` is again uniform on `[0, 1]` whenever `U` is. Such a map is in
general many-to-one: it folds the interval onto itself so that the pieces stack
up to a flat density. This vignette tours the udp families in the package; see
`vignette("aceshuffle")` for a use of shuffles to uncover non-monotonic
dependence.

Throughout we work with a single uniform sample.

```{r data}
set.seed(1)
n <- 1000
U <- runif(n)
```

## A v-transform

A v-transform has a single fulcrum `delta`: values below it are reflected
upward, values above it are carried down, and the two branches are stretched so
that uniformity is preserved. `v2p()` builds a smooth two-parameter version.

```{r vtransform}
T1 <- v2p(delta = 0.4, kappa = 0.8)
plot(T1, embellish = "colour")
```

`udptrans()` applies the map. The transformed sample is still uniform:

```{r vtransform-forward}
V <- udptrans(T1, U)
hist(V, main = "udptrans(T1, U)")
```

Because the map is two-to-one, it has no ordinary inverse. `udpsi()` instead
draws one pre-image at random, in the proportion that keeps the uniform
property running backwards: feed it a uniform `V` and an independent uniform
`Z` and the result is uniform again.

```{r vtransform-inverse}
U2 <- udpsi(T1, V)
hist(U2, main = "udpsi(T1, V)")
```

`U2` is a valid pre-image of `V`, but usually not the `U` we started from. The
probability that stochastic inversion lands back on the original value is
returned by `pcoincide()`.

```{r vtransform-pcoincide}
plot(U, U2, pch = 20, cex = 0.4, asp = 1, xaxs = "i", yaxs = "i")
pcoincide(T1)
```

## A shuffle

A shuffle cuts `[0, 1]` into `m` equal strips, permutes them, and optionally
flips the slope on each. It is a piecewise-linear bijection, so it preserves
uniformity exactly and `udpsi()` recovers the input.

```{r shuffle}
perm <- sample(1:20)
signs <- rbinom(20, 1, 0.5) * 2 - 1
T2 <- shuffle(perm, signs)
plot(T2, embellish = "colour")

hist(udptrans(T2, U), main = "udptrans(T2, U)")
```

## A udp function from a cosine curve

`udpcosine(k)` is the degree-`k` triangle wave: `k` linear pieces of slope
`±k` that fold the interval into `k` layers.

```{r cosine}
T3 <- udpcosine(6)
plot(T3, embellish = "colour")

V <- udptrans(T3, U)
U2 <- udpsi(T3, V)
pcoincide(T3)
```

The `k` pre-images are equally weighted, so `pcoincide()` is `1 / k`.

## A udp function from a zigzag wave

`udpzigzag()` generalizes the cosine construction to unequal-width pieces:
`[0, 1]` is partitioned into pieces of any positive widths, each stretched
into an affine bijection onto the whole of `[0, 1]` and alternating direction
so the map stays continuous. Equal widths recover `udpcosine(k)`; two pieces
recover a `vlinear()` v-transform.

```{r zigzag}
T4 <- udpzigzag(widths = c(3, 1, 2, 4, 1))
plot(T4, embellish = "colour")

V <- udptrans(T4, U)
U2 <- udpsi(T4, V)
pcoincide(T4)
```

Because every piece maps onto the whole of `[0, 1]`, the raw selection weight
`1 / |T'|` at each pre-image is exactly that piece's width -- constant rather
than varying with `v` -- so `pcoincide()` is available as the closed form
`sum(widths^2)` (widths normalized to sum to `1`).

```{r zigzag-pcoincide}
w <- c(3, 1, 2, 4, 1)
sum((w / sum(w))^2)
```

## A udp function from a Legendre polynomial

`udplegendre(j)` composes the shifted Legendre polynomial `L_j` with the
distribution function of `L_j(U)`. It has up to `j` pre-images, weighted by
`1 / |L_j'|`.

```{r legendre}
T5 <- udplegendre(4)
plot(T5, embellish = "colour")

V <- udptrans(T5, U)
U2 <- udpsi(T5, V)
pcoincide(T5)
```

## References

- McNeil, A. J. (2021). Modelling volatile time series with v-transforms and
  copulas. *Risks*, **9**(1), 14.
- McNeil, A. J., Nešlehová, J. G. and Smith, A. D. (2025). Measures and models
  of non-monotonic dependence. arXiv:2512.10828.
