---
title: "Uniform distribution preserving transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uniform distribution preserving transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4,
  fig.align = "center"
)
```

```{r library}
library(udp)
```

A map `T` of the unit interval to itself is **uniform distribution preserving**
(udp) if `T(U)` is again uniform on `[0, 1]` whenever `U` is. Such a map is in
general many-to-one: it folds the interval onto itself so that the pieces stack
up to a flat density. This vignette tours the udp families in the package and
finishes with a use of shuffles to uncover non-monotonic dependence.

Throughout we work with a single uniform sample.

```{r data}
set.seed(1)
n <- 1000
U <- runif(n)
```

## A v-transform

A v-transform has a single fulcrum `delta`: values below it are reflected
upward, values above it are carried down, and the two branches are stretched so
that uniformity is preserved. `v2p()` builds a smooth two-parameter version.

```{r vtransform}
T1 <- v2p(delta = 0.4, kappa = 0.8)
plot(T1)
```

`udptrans()` applies the map. The transformed sample is still uniform:

```{r vtransform-forward}
V <- udptrans(T1, U)
hist(V, main = "udptrans(T1, U)")
```

Because the map is two-to-one, it has no ordinary inverse. `udpsi()` instead
draws one pre-image at random, in the proportion that keeps the uniform
property running backwards: feed it a uniform `V` and an independent uniform
`Z` and the result is uniform again.

```{r vtransform-inverse}
U2 <- udpsi(T1, V)
hist(U2, main = "udpsi(T1, V)")
```

`U2` is a valid pre-image of `V`, but usually not the `U` we started from. The
probability that stochastic inversion lands back on the original value is
returned by `pcoincide()`.

```{r vtransform-pcoincide}
plot(U, U2, pch = 20, cex = 0.4)
pcoincide(T1)
```

## A shuffle

A shuffle cuts `[0, 1]` into `m` equal strips, permutes them, and optionally
flips the slope on each. It is a piecewise-linear bijection, so it preserves
uniformity exactly and `udpsi()` recovers the input.

```{r shuffle}
perm <- sample(1:20)
signs <- rbinom(20, 1, 0.5) * 2 - 1
T2 <- shuffle(perm, signs)
plot(T2)

hist(udptrans(T2, U), main = "udptrans(T2, U)")
```

## A udp function from a cosine curve

`udpcosine(k)` is the degree-`k` triangle wave: `k` linear pieces of slope
`±k` that fold the interval into `k` layers.

```{r cosine}
T3 <- udpcosine(6)
plot(T3)

V <- udptrans(T3, U)
U2 <- udpsi(T3, V)
pcoincide(T3)
```

The `k` pre-images are equally weighted, so `pcoincide()` is `1 / k`.

## A udp function from a Legendre polynomial

`udplegendre(j)` composes the shifted Legendre polynomial `L_j` with the
distribution function of `L_j(U)`. It has up to `j` pre-images, weighted by
`1 / |L_j'|`.

```{r legendre}
T4 <- udplegendre(5)
plot(T4)

V <- udptrans(T4, U)
U2 <- udpsi(T4, V)
pcoincide(T4)
```

## Uncovering non-monotonic dependence with shuffles

Stochastic inversion lets us build a dataset with a deliberately awkward
dependence structure. Start from a strongly dependent copula sample and pull
each margin back through a udp function. This section uses the **copula**
package.

```{r ace-guard, include = FALSE}
knitr::opts_chunk$set(eval = requireNamespace("copula", quietly = TRUE))
```

```{r ace-data}
set.seed(1)
GC <- copula::gumbelCopula(3)
W <- copula::rCopula(n, GC)
X <- cbind(udpsi(T4, W[, 1]), udpsi(T3, W[, 2]))
plot(X, pch = 20, cex = 0.4, xlab = "X1", ylab = "X2")
```

The margins of `X` are uniform and their ordinary correlation is close to zero,
yet they are far from independent. `aceshuffle()` searches, in the manner of
alternating conditional expectations, for a shuffle of each margin that makes
the transformed pair as linearly correlated as possible.

```{r ace-fit}
fit <- aceshuffle(X[, 1], X[, 2], m = 50)
fit$iterations
fit$correlation
```

The fitted shuffles approximate the udp functions we used to build the data.
The fitted pair is unique only up to the joint reflection
`(shuffle1, shuffle2) -> (1 - shuffle1, 1 - shuffle2)`, which leaves the
correlation unchanged; `aceshuffle()` returns the orientation whose shuffles
trend upward.

```{r ace-shapes, fig.show = "hold", fig.width = 4}
u <- seq(0, 1, length.out = 200)

plot(fit$shuffle1)
lines(u, udptrans(T4, u), col = 2, lwd = 2)

plot(fit$shuffle2)
lines(u, udptrans(T3, u), col = 2, lwd = 2)
```

When a generating transform is close to symmetric about the centre of the unit
square (as `udpcosine(6)` is), the trend rule cannot pin the orientation from
that shuffle alone; here the clear upward trend of the `udplegendre(5)` shuffle
fixes the orientation of the pair. If an overlay comes out reflected, compare
against `1 - udptrans(...)` instead.

`fit$V` holds the shuffled data, `cbind(udptrans(shuffle1, X1),
udptrans(shuffle2, X2))`. Its correlation matches the copula it came from.

```{r ace-result}
plot(fit$V, pch = 20, cex = 0.4)
c(fitted = fit$correlation, original = cor(W)[1, 2])
```

## References

- McNeil, A. J. (2021). Modelling volatile time series with v-transforms and
  copulas. *Risks*, **9**(1), 14.
- McNeil, A. J., Nešlehová, J. G. and Smith, A. D. (2025). Measures and models
  of non-monotonic dependence. arXiv:2512.10828.
